Binning Strategies and Related Loss for Large Data

### 2015-04-17 Paper.Rnw
discussion of trade-off of spatial precision for frequecy information added to Section 4.2 (also, spell check)

\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely. A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell) \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies. This is analogous to the fundimental problem for overplotted standard scatterplots where frequency is masked by other points. 
\end{itemize}
}

@@

\ktm{The net spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.  If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.  However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard devations. }

@@

@@

\ktm{It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units will minimize the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimesions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}). }

@@

In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable choice for lowering spatial loss for symmetric bivariate data at a coarse resolution while using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
%\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss. it's a bit confusing with the two minima floating around. It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
\ktm{Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm. Note that for simplicity, only origin offsets for with equal offset in each dimension are included in this simulation. Since the data resolution is two units in each dimension, we will pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees. The round glyphs indicade the origin offset where the net loss reached an absolute minimum. For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.

@@

\hh{XXX Let's try a first picture with a continuous scale and three discrete binned frequencies of four, seven, and ten bins. This should drive home the message that Healey is right, and we can't distinguish more than seven shades anyway, i.e. we don't see a difference between the continuously shaded plot and the one with discrete shades. This re-enforces our approach to treat color as a categorical variable, because we preceive it that way (otherwise why would we?)}

@@

\ktm{We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading. It is recommended to use less than ten seperate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins. Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.  }

>>>>

\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely. A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell) \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies. This is analogous to the fundamental problem for over-plotted standard scatterplots where frequency is masked by other points. 
\end{itemize}
}

@@

\ktm{The net spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded.  If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.  However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations. }

@@

\ktm{It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates. The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from overplotting, which can be immense for large data sets. When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location. This implicit loss of frequency information is then exacerbated when points are rendered as a circle with a radius large enough to also partially cover nearby points. In binned scatterplots, we have traded the exact locations for exact frequencies within bins. }

@@

\ktm{It can be shown analytically that an origin offset of ($\alpha_x/2$,$\alpha_y/2$) units will minimize the net spatial loss in the situation with the following three properties: (i) data are recorded to a resolution of $\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension, (ii) points are symmetric distributed within rectangular bins, (iii) the bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$, respectively (see proof in Appendix~\ref{proof:offset}). }

@@

In practice the ($\alpha_x/2$,$\alpha_y/2$) origin offset is found to be a reasonable choice for lowering spatial loss for symmetric bivariate data at a coarse resolution while using bin dimensions that are integer multiples of $\alpha_x$ and $\alpha_y$.
%\hh{In the next paragraph you are refering to the minimal value (as in minimal data value) and you are minimizing spatial loss. it's a bit confusing with the two minima floating around. It might be better to properly define the offset, and then refer to offset values rather than origins below the minimal value}
\ktm{Figure~\ref{fig:OriginOffsetSimulations} shows how the net spatial loss changes as the origin offset is changed while binning the coarse simulated data using the standard rectangular binning algorithm. Note that for simplicity, only origin offsets for with equal offset in each dimension are included in this simulation. Since the data resolution is two units in each dimension, we will pay special attention to an origin offset of (1,1) in order to assess how well the theoretically optimal origin offset at ($\alpha_x/2$,$\alpha_y/2$) works for data that violate the assumptions to differing degrees. The round glyphs indicate the origin offset where the net loss reached an absolute minimum. For the two symmetrically distributed data sets, normal and uniform, the origin offset of one unit was found to either minimize the net spatial loss or achieve a local minimum very near to the overall minimum (within a 0.2\% increase from the minimum spatial loss) for each considered bin size.} %\hh{XXX - no! the value at -1 is close, but the dots are in the actual minimal values. This is only a single example - the expected minimum might still be in -1. }
For the bivariately skewed exponential data, the 1 unit origin offset minimized net loss for the smaller intervals but was not optimal for the largest intervals; 2.5\% and 7\% increase the minimum spatial loss for bins of size eight and ten, respectively.

@@

\hh{XXX Let's try a first picture with a continuous scale and three discrete binned frequencies of four, seven, and ten bins. This should drive home the message that Healey is right, and we can't distinguish more than seven shades anyway, i.e. we don't see a difference between the continuously shaded plot and the one with discrete shades. This re-enforces our approach to treat color as a categorical variable, because we perceive it that way (otherwise why would we?)}

@@

\ktm{We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading. It is recommended to use less than ten separate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins. Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.  }
